---
title: "Storm Data Analysis"
author: "Casey Thomas"
date: "March 15, 2015"
output: 
  html_document:
    keep_md: true
---

## Snyopsis

describes and summarizes your analysis in at most 10 complete sentences.

## Data Processing
```{r setoptions, echo=TRUE, cache=TRUE}
library('dplyr')
library('ggplot2')
library('scales')
library('stringr')
library('reshape2')
library('magrittr')

setwd('~/git/data-science/RepData_PeerAssessment2/')

data <- read.csv('data/repdata-data-StormData.csv', as.is = TRUE)
```

Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

 
```{r, cache=FALSE}
harmful_to_pop_by_type <- data %>%
  group_by(EVTYPE) %>%
  filter(FATALITIES > 0 | INJURIES > 0) %>%
  summarize(total_fatalities = sum(FATALITIES),
            total_injuries = sum(INJURIES)) %>%
  arrange(desc(total_injuries),
          desc(total_fatalities),
          EVTYPE)

harmful_to_pop_by_type

harmful_to_pop_by_type %>%
  head(25) %>%
  melt(id.vars = "EVTYPE") %>%
  ggplot(aes(x = EVTYPE, y = value, fill = variable)) +
    geom_bar(stat = "Identity") +
    coord_flip()
```

Across the United States, which types of events have the greatest economic consequences?

```{r, cache=FALSE}
econ_damage_by_type <- data %>%
  group_by(EVTYPE) %>%
  filter(PROPDMG > 0 | CROPDMG > 0) %>%
  summarize(total_propdmg = sum(PROPDMG),
            total_cropdmg = sum(CROPDMG)) %>%
  arrange(desc(total_propdmg),
          desc(total_cropdmg),
          EVTYPE)

econ_damage_by_type %>%
  head(25) %>%
  melt(id.vars = "EVTYPE") %>%
  ggplot(aes(x = EVTYPE, y = value, fill = variable)) +
    geom_bar(stat = "Identity") +
    coord_flip()
```

## Results

Tornados are by far the most the natural disaster that is harmful with to people and property.

